Cost of Living in Toronto

### 2024-09-24 paper/paper.qmd
added citations in references.bib

```{r}
#| include: false
#| warning: false
#| message: false
#install.packages('opendatatoronto')
#install.packages("dplyr")
#install.packages('tidyverse')
#install.packages('sf')
#remove.packages('knitr')
#install.packages('xfun')
#install.packages('knitr')

# RUN THIS IN THE TERMINAL IF THERE'S AN ERROR IN EXPORTING TO PDF
# Source: https://stackoverflow.com/questions/75906430/why-cant-i-render-any-quarto-documents-into-pdfs
# quarto install tool tinytex

library(here)
library(knitr)
library(opendatatoronto)
library(tidyverse)
library(dplyr)
library(sf)

@@

While Toronto may seem like a perfect place to live, residents have said this to be far from the truth, with many struggling financially to meet ends, especially for those from a low income background. For example, according to the Daily Bread Food Bank, `1/10` people in Toronto relied on food banks in 2023, a huge increase from `1/20` from 2022 (CITATION HERE).

According to the City of Toronto, the threshold for `LIM_AT` (threshold to be considered low income) was `$26,503` for 1 person household, `$37,480` for two person household and `$53,005` for a four person household (CITATION, CITY OF TORONTO). `13.2%` of Toronto's population fell under this category of low-income. Furthermore, Toronto continues to have a higher rate of low income than any other regions in the GTHA at `8.09%` (CITATION HERE), highlighting an issue of income discrepancy.

@@

This report uses the "About Cost of living in Toronto for low-income households" (CITATION HERE) dataset, which is published publicly by the Social Development, Finance & Administration of Toronto on City of Toronto's Open Data Portal. This data contains information on income such as employment income, GST/HST credits, pension benefits and expenses such as rent, food, transportation costs for 11 different scenarios of a low-income household. The 11 different scenarios of low-income household were manually selected by the City of Toronto to reflect different household characteristics that can increase the reliance on social assistance, such as type of income, single-parent household, etc. A detailed explanation of this 11 household scenarios can be found in the Appendix. The dataset is also within a monthly time period, reflecting monthly income and expenses.

The dataset was downloaded and analyzed using the R Programming Language (R Core Team 2022), opendatatoronto (CITATION HERE), knitr (CITATION HERE), tidyverse (CITATION HERE), here (CITATION), dplyr (CITATION) and sf (CITATION HERE).

@@

The dataset published by `Social Development, Finance & Administration of Toronto` was last updated on August 2024 on an annual basis. However, the dataset uses price estimates from July 2022 - July 2023 for its income and expenses of the families. The variables chosen for this analysis were variables for the 11 households, for each entry of the income and expense types. See Table 1 for an overview. It is noted that expenses and Households 4 to 11 are not visible in the preview, and expenses are not visible due to the long list of income that is above it.

@@

`Table 2` shows that income from employment was the major income source for all the 11 households in Toronto by a large margin, at approximately $1603 per month. This number also falls shy below the "low-income" threshold set by the City of Toronto, which is approximately $2208.5 (CITY OF TORONTO CITATION).

@@

Without context to this information, determining the efficiency of spending of the families is impossible. However, looking at the results suggest some inefficiencies in spending. For example, in `Figure 3`, the average rent was $3146 per month for the 11 families, a big difference from the average rent for a one-bedroom condo of $2,529 in Toronto for Q2 of 2023 (Toronto Regional Estate Board, 2024). Expenses such as this could suggest that expenses could be correlated with poor financial decisions, however more research must be conducted to draw those conclusions.

@@

In summary, expenses exceeded income for most of the 11 household scenarios generated by the `City of Toronto`, demonstrating low-income families may not be able to make ends meet financially. Furthermore, rent made up a significant portion of the families' expenses, guiding policymakers to focus on this area. Further research on consumer spending behavior could better explain the expenses behind the families.

@@

Household 1: 2 adults (male and female ages 31-50) with no income, 2 children (unable to contribute to income).
Household 2: 2 adults (male and female ages 31-50) with minimum wage pay ($15.00/hr), 2 children (unable to contribute to income).
Household 3: 2 adults (male and female ages 31-50) with median income, 2 children (unable to contribute to income).
Household 4: 1 adult (female age 31-50) with no income, 2 children (unable to contribute to income).
Household 5: 1 adult (male age 31-50) with no income.
Household 6: 1 adult (male age 31-50) under `Ontario Disability Support Program`, 2 children (unable to contribute to income).
Household 7: 1 adult (female age 70+) under Old Age Security/Guaranteed Annual Income Supplement.
Household 8: 2 adults (male and female age 31-50) under Ontario Disability Support Program.
Household 9: 1 adult (female age 19-30) under Ontario Disability Support Program.
Household 10: 1 adult (female age 31-50) with minimum wage pay ($15.00/hr) with 2 children (unable to contribute to income).
Household 11: 1 adult (female age 31-50) with minimum wage pay ($15.00/hr).

@@

https://www.toronto.ca/wp-content/uploads/2022/07/9877-City-Planning-2021-Census-Backgrounder-Families-Hhlds-Marital-Status-Income.pdf

https://www.dailybread.ca/wp-content/uploads/2023/11/DB-WhosHungryReport-2023-Digital.pdf

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
#install.packages('opendatatoronto')
#install.packages("dplyr")
#install.packages('tidyverse')
#install.packages('sf')
#remove.packages('knitr')S
#install.packages('knitr')
#install.packages('ggplot2)

# RUN THIS IN THE TERMINAL IF THERE'S AN ERROR IN EXPORTING TO PDF
# Source: https://stackoverflow.com/questions/75906430/why-cant-i-render-any-quarto-documents-into-pdfs
# quarto install tool tinytex
library(ggplot2)
library(here)
library(knitr)
library(opendatatoronto)
library(tidyverse)
library(dplyr)
library(sf)

@@

While Toronto may seem like a perfect place to live, residents have said this to be far from the truth, with many struggling financially to meet ends, especially for those from a low income background. For example, according to the Daily Bread Food Bank, `1/10` people in Toronto relied on food banks in 2023, a huge increase from `1/20` from 2022 [@citewhoshungry].

According to the City of Toronto, the threshold for `LIM_AT` (threshold to be considered low income) was `$26,503` for 1 person household, `$37,480` for two person household and `$53,005` for a four person household [@citecityoftoronto]. `13.2%` of Toronto's population fell under this category of low-income. Furthermore, Toronto continues to have a higher rate of low income than any other regions in the GTHA at `8.09%` [@citecityoftoronto], highlighting an issue of income discrepancy.

@@

This report uses the "About Cost of living in Toronto for low-income households" [@citecityoftoronto] dataset, which is published publicly by the Social Development, Finance & Administration of Toronto on City of Toronto's Open Data Portal. This data contains information on income such as employment income, GST/HST credits, pension benefits and expenses such as rent, food, transportation costs for 11 different scenarios of a low-income household. The 11 different scenarios of low-income household were manually selected by the City of Toronto to reflect different household characteristics that can increase the reliance on social assistance, such as type of income, single-parent household, etc. A detailed explanation of this 11 household scenarios can be found in the Appendix. The dataset is also within a monthly time period, reflecting monthly income and expenses.

The dataset was downloaded and analyzed using the `R Programming Language` [@citeR], `opendatatoronto` [@citeopendatatoronto], `knitr` [@citeknitr], `tidyverse` [@citetidyverse], `here` [@citehere], `dplyr` [@citedplyr] and `sf` [@citesf].

@@

The dataset published by `Social Development, Finance & Administration of Toronto` was last updated on August 2024 on an annual basis. However, the dataset uses price estimates from July 2022 - July 2023 for its income and expenses of the families. The variables chosen for this analysis were variables for the 11 households, for each entry of the income and expense types. See Table 1 for an overview. It is noted that expenses are not visible in `Table 1` due to the long list of income that occupies the first rows.

@@

`Table 2` shows that income from employment was the major income source for all the 11 households in Toronto by a large margin, at approximately $1603 per month. This number also falls shy below the "low-income" threshold set by the City of Toronto, which is approximately $2208.5 [@citedcityoftoronto].

@@

Without context to this information, determining the efficiency of spending of the families is impossible. However, looking at the results suggest some inefficiencies in spending. For example, in `Figure 3`, the average rent was $3146 per month for the 11 families, a big difference from the average rent for a one-bedroom condo of $2,529 in Toronto for Q2 of 2023 [@citerentalmarket]. Expenses such as this could suggest that expenses could be correlated with poor financial decisions, however more research must be conducted to draw those conclusions.

@@

@@

`Household 1`: 2 adults (male and female ages 31-50) with no income, 2 children (unable to contribute to income).

`Household 2`: 2 adults (male and female ages 31-50) with minimum wage pay ($15.00/hr), 2 children (unable to contribute to income).

`Household 3`: 2 adults (male and female ages 31-50) with median income, 2 children (unable to contribute to income).

`Household 4`: 1 adult (female age 31-50) with no income, 2 children (unable to contribute to income).

`Household 5`: 1 adult (male age 31-50) with no income.

`Household 6`: 1 adult (male age 31-50) under `Ontario Disability Support Program`, 2 children (unable to contribute to income).

`Household 7`: 1 adult (female age 70+) under Old Age Security/Guaranteed Annual Income Supplement.

`Household 8`: 2 adults (male and female age 31-50) under Ontario Disability Support Program.

`Household 9`: 1 adult (female age 19-30) under Ontario Disability Support Program.

`Household 10`: 1 adult (female age 31-50) with minimum wage pay ($15.00/hr) with 2 children (unable to contribute to income).

`Household 11`: 1 adult (female age 31-50) with minimum wage pay ($15.00/hr).

@@
